---
title: "Reproducible Research: Peer Assessment 1"
author: "Aleksander Petrovskii aka petr0vsk"
date: "March 21, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(tidyr)
library(dplyr)
library(stringr)
library(ggplot2)
```
## 1. Loading and preprocessing the data

```{r echo = TRUE}

project.dir <- "/home/petr0vsk/Project3"
stopifnot( dir.exists(file.path(project.dir))  )
setwd(file.path(project.dir))
steps.raw <- read.csv("activity.csv",  header = TRUE) 
str(steps.raw)
```

## 2. What is mean total number of steps taken per day?
1. Make a histogram of the total number of steps taken each day
Calculate total number of steps taken each day
```{r echo = TRUE}

sum.steps.per.day.withNA <- steps.raw %>%
    group_by(date) %>%
    summarise_each(funs( sum(steps, na.rm = T) ), steps = steps) %>%
as.data.frame()
```
Histogram of the total number of steps taken each day
```{r echo=TRUE}
hist(sum.steps.per.day.withNA$steps, main = "Total number of steps taken each day with NA", 
     breaks = 61,
     xlab = "Steps per day", ylab = "Frequency",
     col="black",
     border="white")
box(bty="l")
grid(nx=NA,ny=NULL,lty=1,lwd=1,col="gray")
rug(sum.steps.per.day.withNA$steps)
```
Calculate and report the mean and median total number of steps taken per day
```{r echo=TRUE}
mn.NA <- round(mean(sum.steps.per.day.withNA$steps, na.rm = T),2)
md.NA <- median(sum.steps.per.day.withNA$steps, na.rm = TRUE)
print(paste0("mean.with.NA = ", mn.NA))
print(paste0("median.with.NA = ", md.NA))
```
## 3. What is the average daily activity pattern?
```{r echo=TRUE}
daily.activity.average.with.NA <- steps.raw %>%
    group_by(interval) %>%
summarise_each(funs(mean(steps, na.rm = TRUE)), steps = steps) 
```

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
and the average number of steps taken, averaged across all days (y-axis)
```{r echo=TRUE}
brake.vec <- as.vector(seq(1,288,by=12))
ggplot(daily.activity.average.with.NA, aes(x=interval, y=steps)) +
    geom_line(colour = "darkblue") +                                                                            geom_point(size=0.7) +
    scale_x_continuous(name = "Time interval (by 5-minyt step)", limits = c(0,tail(daily.activity.average.with.NA$interval, n=1)), breaks = daily.activity.average.with.NA$interval[brake.vec])  +
    scale_y_continuous(name = "Time series plot of the average number of steps taken", limits = c(0,max(daily.activity.average.with.NA$steps))) +
    ggtitle("Mean total number of steps taken per day") +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x = element_text(angle=30, hjust=1, vjust=1)) +
stat_smooth(colour="green", method = 'loess', na.rm=TRUE) 
```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE}
max.steps <- daily.activity.average.with.NA[which.max(daily.activity.average.with.NA$steps),]
print(paste0("5-minute interval, contains the maximum number of steps is", max.steps))
```
## 4. Imputing missing values
All of the missing values are replased with mean value for that 5-minute interval
Calculate and report the total number of missing values in the dataset
```{r echo=TRUE}
print(paste0("Total number of missing values in the datase = ", sum(is.na(steps.raw$steps)) ))
```
Create a new dataset that is equal to the original dataset but with the missing data filled in
replase NA with median of interval
```{r echo=TRUE}
steps.raw.clear <- steps.raw
for(i in 1:length(steps.raw.clear$steps))   {
    if( is.na(steps.raw.clear$steps[i]) )        {
        steps.raw.clear$steps[i]<- filter(daily.activity.average.with.NA, interval==steps.raw.clear$interval[i])%>%
                  select(steps) %>%
                  round(2)
    }#if..
}#for..
steps.raw.clear$steps <- unlist(steps.raw.clear$steps)
```
## 5. Are there differences in activity patterns between weekdays and weekends?
What is sum of total number of steps taken per day?
```{r echo=TRUE}
sum.steps.per.day.withoutNA <- steps.raw.clear %>%
    group_by(date) %>%
    summarise_each(funs( sum(steps) ), steps = steps) 
mn <- round(mean(sum.steps.per.day.withoutNA$steps),2)
md <- median(sum.steps.per.day.withoutNA$steps)
```
Histogram of the total number of steps taken each day without NA
```{r echo=TRUE}
par(mfrow= c(2,1))
hist(sum.steps.per.day.withoutNA$steps, main = "Total number of steps after missing values are imputed", 
     breaks = 61,
     xlab = "Steps per day", ylab = "Frequency",
     col="black",
     border="white")
box(bty="l")
grid(nx=NA,ny=NULL,lty=1,lwd=1,col="gray")
rug(sum.steps.per.day.withoutNA$steps)

hist(sum.steps.per.day.withNA$steps, main = "Total number of steps taken each day with NA", 
     breaks = 61,
     xlab = "Steps per day", ylab = "Frequency",
     col="black",
     border="white")
box(bty="l")
grid(nx=NA,ny=NULL,lty=1,lwd=1,col="gray")
rug(sum.steps.per.day.withNA$steps)
```
Print mean and median
```{r echo=TRUE}
print(paste0("mean.with.NA = ", mn.NA))
print(paste0("median.with.NA = ", md.NA))
print(paste0("after missing values are imputed mean = ", mn))
print(paste0("after missing values are imputed median = ", md))

```
Find the day of the week for each measurement in the dataset
```{r echo=TRUE}
steps.raw.clear$date <- as.Date(steps.raw.clear$date)
for(i in 1:length(steps.raw.clear$steps))   {
      if( weekdays(steps.raw.clear$date[i]) %in% c("Saturday", "Sunday")  )   {
        steps.raw.clear$weekday[i] <-  "weekend"     
                    }#if..
        else {
            steps.raw.clear$weekday[i] <-  "weekday"
        } 
}#for..
```
Panel plot comparing the average number of steps taken per 5-minute interval 
across weekdays and weekends 
```{r echo=TRUE}
mean.steps.raw.clear <- steps.raw.clear %>%
    group_by(interval,weekday) %>%
    summarise_each(funs( mean(steps)), steps = steps, weekday = weekday) 
ggplot(mean.steps.raw.clear, aes(x=interval, y=steps)) +
    geom_line(colour = "darkblue") +  
    facet_grid(weekday ~ .) + 
    geom_point(size=0.7) +
    scale_x_continuous(name = "Time interval (by 5-minyt step)", limits = c(0,tail(mean.steps.raw.clear$interval, n=1)), breaks = daily.activity.average.with.NA$interval[brake.vec])  +
    scale_y_continuous(name = "Time series plot of the average number of steps taken", limits = c(0,max(mean.steps.raw.clear$steps))) +
    ggtitle("Mean total number of steps taken per day") +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x = element_text(angle=30, hjust=1, vjust=1)) +
stat_smooth(colour="green", method = 'loess', na.rm=TRUE) 
```
